Stochastic Gradient Descent

### 2018-04-25 sgd/sgd.tex
final presentation

\begin{frame}
	\frametitle{Optimization is fundamental to statistical modeling}
	Suppose that we have a typical supervised classification problem
	\begin{itemize}
		\item \small Non-parametric: no assumptions about distribution of data
		\item Feature vector $\mathbf{X}_i$, label $Y_i$
		\item Want to find best prediction function $f_w^*$ from class $\mathcal{F}$
		\item Optimization: pick weights $\hat{\bm{w}}$ that minimize
			empirical risk according to some convex
			loss function $L(f_w(x), y)$ \\~\\
	\end{itemize}

	Our lack of assumptions takes away some familiar tools for finding $\hat{\mathbf{w}}$
	\begin{itemize}
		\item \small Cannot analytically identify $\hat{\mathbf{w}}$
			(e.g. in OLS $=(X'X)^{-1}X'Y$))
		\item \small Newton-Raphson requires 2nd derivative of $L$,
			which might be a pain (unlike in, for example, GLM)\\~\\
	\end{itemize}

	But we still know that if $L$ is convex, there is a unique global
	minimum, and the gradient at that point must be 0
\end{frame}

@@

	$\mathbf{w}_{t+1} = \mathbf{w}_t - \gamma \frac{1}{n}\sum_{i=1}^n
	\nabla_w L(f_w(\mathbf{X}_i), y_i)$ \\~\\

@@

	Theoretical guarantees on speed of convergence ($\rho :=$ size of error):
	\begin{itemize}
		\item Version presented here: $ - \log \rho \sim t$
		\item More optimized version: $ - \log \log \rho \sim t$ \\~\\
	\end{itemize}

@@

	$$
	\textcolor{red}{\mathbf{w}_{t+1}} = \mathbf{w}_t - \gamma \frac{1}{n}
	\textcolor{blue}{\sum_{i=1}^n}
	\nabla_w L(f_w(\mathbf{X}_i), y_i)
	$$\\~\\

	This becomes computationally intractable as $n$ grows large 
	\begin{itemize}
		\item \small If we have $n$ = 10 million, we have to evaluate the
			gradient 10 million times for every step (and the
			algorithm may take many steps to converge)\\~\\
		\item \small May take hours or days to converge \\~\\
	\end{itemize}
\end{frame}

@@

	$$
	{\mathbf{w}_{t+1}} = \mathbf{w}_t - \gamma \nabla_w
	L(f_w(\mathbf{X}_{\textcolor{red}{i}}), y_{\textcolor{red}{i})}
	$$ \\~\\

@@

	\begin{figure}
	\centering
	\includegraphics[scale=0.55]{comparison}
	\end{figure}

@@

\begin{frame}
	\frametitle{SGD is widely used in industry}
	If a Silicon Valley press release uses any of the following phrases...
	\begin{itemize}
		\item \small ``Neural networks''
		\item ``Machine learning''
		\item ``AI'' 
	\end{itemize}

>>>>

\begin{frame}
	\frametitle{Estimating model coefficients requires optimization}
	We typically fit statistical models to maximize some measure of goodness
	(e.g., likelihood) or minimize some measure of badness (e.g., risk)
	\\~\\

	Parametric assumptions can make this optimization very nice \\~\\

	OLS has a closed form solution: $\hat{\bm{\beta}} =
	(\bm{X}'\bm{X})\bm{X}'\bm{Y}$ \\~\\

	GLM often uses Newton's method
	\begin{itemize}
		\item \small Knowing the PDF makes finding the
			Hessian (2nd order derivatives) not too onerous \\~\\
	\end{itemize}

\begin{frame}
	\frametitle{Non-parametric \& high-dimensional settings}
	Suppose that we have a typical supervised classification problem
	\begin{itemize}
		\item \small Non-parametric: no assumptions about distribution of data
		\item Feature vector $\mathbf{X}_i$, label $Y_i$
		\item Want to find best prediction function $f^*(\bm{w}; \bm{X})$ from
			hypothesis class $\mathcal{F}$
		\item Optimization: pick weights $\hat{\bm{w}}$ that minimize
			empirical risk according to some convex
			loss function \\~\\
	\end{itemize}

	Some familiar tools for finding $\hat{\mathbf{w}}$ no longer work well
	\\~\\

	But we still know that if $L$ is convex, there is a unique global
	minimum, and the gradient at that point must be 0 \\~\\

	Gradient descent is a numerical strategy to search for that point
\end{frame}

@@

	$\mathbf{w}^{(t+1)} = \mathbf{w}^{(t)} - \gamma \frac{1}{n}\sum_{i=1}^n
	\nabla_w L(f(\bm{w}^{(t)}; \mathbf{X}_i), Y_i)$ \\~\\

@@

@@

	$$
	\textcolor{red}{\mathbf{w}^{(t+1)}} = \mathbf{w}^{(t)} - \gamma \frac{1}{n}
	\textcolor{blue}{\sum_{i=1}^n}
	\nabla_w L(f(\bm{w}^{(t)}; \mathbf{X}_i), Y_i)
	$$\\~\\

	This becomes computationally intractable as $n$ grows large 
	\begin{itemize}
		\item \small If $n$ = 10 million, have to evaluate 
			gradient 10 million times every step
	\end{itemize}
\end{frame}

@@

	$$
	{\mathbf{w}^{(t+1)}} = \mathbf{w}^{(t)} - \gamma \nabla_w
	L(f(\bm{w}^{(w)}; \mathbf{X}_{\textcolor{red}{i}}), Y_{\textcolor{red}{i})}
	$$ \\~\\

@@

	\begin{figure}
		\caption{Fitting OLS coefficients, $n=10$ million}
	\centering
	\includegraphics[scale=0.35]{speed}
	\end{figure}

@@

\begin{frame}
	\frametitle{SGD is very popular in academia and industry}
	Many, many statistical learning models use SGD to fit weights \\~\\

	If a Silicon Valley press release uses any of the following phrases...
	\begin{itemize}
		\item \small ``Neural networks''
		\item ``Machine learning''
		\item ``AI'' 
	\end{itemize}
